---
title: "IDS investigation worksheet"
author: "by Team-Name: User1, User2, User3, User4 & User5"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(readODS)
library(janitor) #for changing row names

library(ggplot2)
library(RColorBrewer)

```


```{r load-data}
Ethnic_PD <- read_ods("data/Population_30June2023_Annual.ods", sheet = 14)

```


```{r ethnicity-analysis}

male_allNat_Ethnic_PD <- Ethnic_PD %>%
  slice(.,34:40) %>% #ignores "all nationalities" since it is redundant
  select(!2:9)

male_allNat_Ethnic_PD <- as.data.frame(t(male_allNat_Ethnic_PD)) %>%
  row_to_names(row_number = 1) %>%
  rename(
    
    `Asian people` = "Asian/ Asian British(2)",
    Other = "Other ethnic group(2)",
    `Not recorded` = "Ethnic group not recorded",
    `Black people` = "Black/ African/ Caribbean/ Black British",
    `Mixed or Multiple` = "Mixed/ Multiple ethnic groups",
    `White people` = "White"
  )

rownames(male_allNat_Ethnic_PD) <- NULL
male_allNat_Ethnic_PD$Year <- c(2009:2023)


male_allNat_Ethnic_PD %>%
  #setting up data for the plot
  pivot_longer(!Year, names_to = "Ethnicity", values_to = "count") %>%
  group_by(Year) %>%
  
  #plotting
  ggplot(
    aes(
      x = Year,
      y = as.numeric(count)
    )
  ) +
  geom_bar(
    aes(
      fill = Ethnicity,
      group = Ethnicity
    ),
    stat = "identity",
  ) +
  facet_wrap(~Ethnicity,
             scales = "free",
             nrow = 2,
             ncol = 4) +
  
  #styling
  theme_light() +
  theme(
    legend.position = "bottom"
  ) +
  labs(
    x = "Year",
    y = "Number of people",
    title = "Prison population in England and Wales, grouped by ethnicity, from year 2009 to 2023",
  )
```

```{r title}

```