---
title: "IDS investigation worksheet"
author: "by Team-Name: User1, User2, User3, User4 & User5"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(ggplot2)

#PACKAGES YOU MIGHT NEED TO INSTALL
library(readODS)
library(janitor) #for changing row names
library(RColorBrewer)
library(ggforce)

```


```{r load-data-and-func}
Ethnic_PD <- read_ods("data/Population_30June2023_Annual.ods", sheet = 14)

#wrapper function for generating title of plots
wrapper <- function(x, ...) 
{
  paste(strwrap(x, ...), collapse = "\n")
}

#normalises data to range from 0 to 1
normalise_01 <- function(x)
{
  (x-min(x))/(max(x)-min(x))
}

```


```{r ethnicity-plot1}

male_allNat_Ethnic_PD <- Ethnic_PD %>%
  slice(.,34:40) %>% #ignores "all nationalities" since it is redundant
  select(!2:9)

male_allNat_Ethnic_PD <- as.data.frame(t(male_allNat_Ethnic_PD)) %>%
  select(1,2,5,3,4,6,7) %>%
  #Using select() to reorder columns such that "Other", "not stated" and "not recorded" are last.
  #This is because they do not classify as an ethnicity (i.e. a mix of all possible ethnicity), hence they don't follow the same trend as defined ethnicity
  row_to_names(row_number = 1) %>%
  rename(
    `Asian people` = "Asian/ Asian British(2)",
    Other = "Other ethnic group(2)",
    `Not recorded` = "Ethnic group not recorded",
    `Black people` = "Black/ African/ Caribbean/ Black British",
    `Mixed or Multiple` = "Mixed/ Multiple ethnic groups",
    `White people` = "White"
  ) %>%
  #converts all columns to numeric
  mutate_all(~ as.numeric(.))

rownames(male_allNat_Ethnic_PD) <- NULL
male_allNat_Ethnic_PD$Year <- c(2009:2023)



### STARTS PLOTTING PRECEDURES

male_allNat_Ethnic_PD %>%
  #setting up data for the plot
  pivot_longer(!Year, names_to = "Ethnicity", values_to = "count") %>%
  group_by(Ethnicity) %>%
  
  #adds percentage
  mutate(group_p = normalise_01(count)) %>%
  
  #plotting
  ggplot(
    aes(
      x = Year,
      y = count,
      fill = group_p
    )
  ) +
  geom_bar(
    aes(
      group = Ethnicity
    ),
    stat = "identity",
  ) +
  facet_wrap(~Ethnicity,
             scales = "free",
             shrink = FALSE,
             nrow = 3,
             ncol = 3) +
  
  #styling
  theme_light() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    x = "Year",
    y = "Number of people",
    title = wrapper("Prison population in England and Wales, grouped by ethnicity, from year 2009 to 2023", 60),
    fill = wrapper("normalised indicator of max and min within Ethnicity group",40)
  ) +
  # scale_fill_gradient(low='red', high='forestgreen', breaks = c(0,1))
  scale_fill_gradient2(low='violetred3', mid='grey', high='chartreuse3',
                       midpoint = 0.5,
                         breaks = c(0,0.5,1),
                       labels = c("min","","max")
                      )
```

```{r ethnicity-plot2}
  #plotting with different viz, namely difference between years
  
  male_allNat_Ethnic_PD %>%
  pivot_longer(!Year, names_to = "Ethnicity", values_to = "count") %>%
  group_by(Ethnicity) %>%
  
  mutate(group_p = normalise_01(c( 0, diff(count,lag = 1)) ) ) %>%
  
  ggplot(
    aes(
      x = Year,
      y = count,
      fill = group_p
    )
  ) +
  geom_bar(
    aes(
      group = Ethnicity
    ),
    stat = "identity",
  ) +
  facet_wrap(~Ethnicity,
             scales = "free",
             shrink = FALSE,
             nrow = 3,
             ncol = 3) +
  
  #styling
  theme_light() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    x = "Year",
    y = "Number of people",
    title = wrapper("Prison population in England and Wales, grouped by ethnicity, from year 2009 to 2023", 60),
    fill = wrapper("normalised indicator of difference from previous year within Ethnicity group",40)
  ) +
  scale_fill_gradient2(low='violetred3', mid='grey', high='chartreuse3',
                       midpoint = 0.5,
                         breaks = c(0,0.5,1),
                       labels = c("decrease","","increase")
                      )
```

```{r title}

```