---
title: 'IDS investigation worksheet'
author: 'by GOMAKASHI: Sky, Hanna, Maggie, Marcel, Siddhi, Vivi'
date: '`r Sys.Date()`'
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE, echo = FALSE}
library(tidyverse)
library(ggplot2)
library(tidymodels)

#PACKAGES YOU MIGHT NEED TO INSTALL
library(readODS)
library(janitor) #for changing row names
library(RColorBrewer)
library(forecast)
library("zoo")
#removed zoo because it is under forecast
```


```{r load-data-and-func}
Ethnic_PD <- read_ods('data/Population_30June2023_Annual.ods', sheet = 14)
A1_7 <- read_ods('data/Population_30June2023_Annual.ods', sheet = 12)
TypeOfCrime <- read_ods('data/Population_30June2023_Annual.ods', sheet = 10)
Custody_PD <- read_ods('data/Population_30June2023_Annual.ods', sheet = 2)
A1_5i <- read_ods('data/Population_30June2023_Annual.ods', sheet = 9)

#wrapper function for generating title of plots
wrapper <- function(x, ...) 
{
  paste(strwrap(x, ...), collapse = '\n')
}

#normalises data to range from 0 to 1
normalise_01 <- function(x)
{
  if (min(x) < 0 )
  {
    x/max(abs(x))
  } else {
    (x-min(x))/(max(x)-min(x))
  }
}

```

```{r custody-data_wrang}
#use facet grid: (<Male/Female>,<Age group>, <custody type>)
offset <- 2

Custody_PD_sliced <- Custody_PD %>%
  row_to_names(row_number = 1) %>%
    rename("Custody" = 1) %>%
  rename("30-6-09" = 10) %>%
  select(!9) %>%
  #removing rows we don't want
  slice(-c(1:(118-offset),(185-offset):(212-offset),(279-offset):nrow(.))) %>% #looks at male and female separately, gets 6 blocks (3 age groups for 2 genders, ignores blocks for total)
  slice(-seq.int(1,5*22+1,by=22)) %>% #removes columns containing totals per category
  cbind(
    #Manually renames age and gender category
    Gender = rep(c("Male","Female"), rep(63,2)),
    Age = rep(c("Adult","18-20", "15-17"), rep(21,3))
  ) %>%
  replace(., . == "-", NA) %>%
  pivot_longer(!c("Custody", "Gender", "Age"), names_to = "Date", values_to = "Count") %>%
  mutate(
    Custody = case_when(
      .default = as.character(Custody),
      # Custody == "Remand" ~ "",
      # Custody == "Untried" ~ "",
      # Custody == "Convicted unsentenced" ~ "",
      # Custody == "Sentenced" ~ "",
      # Custody == "Fine defaulter" ~ "",
      Custody == "Less than or equal to 6 months" ~ "0.5 yrs",
      Custody == "Greater than 6 months to less than 12 months" ~ "0.5-1 yrs",
      Custody == "12 months to less than 4 years" ~ "1-4 yrs",
      Custody == "12 months to less than 2 years" ~ "1-2 yrs",
      Custody == "2 years to less than 4 years" ~ "2-4 yrs",
      Custody == "4 years or more (excluding indeterminate sentences)" ~ ">4.0 yrs",
      Custody == "4 years to less than 5 years" ~ "4-5 yrs",
      Custody == "5 years to less than 7 years" ~ "5-7 yrs",
      Custody == "7 years to less than 10 years" ~ "7-10 yrs",
      Custody == "10 years to less than 14 years" ~ "10-14 yrs",
      Custody == "14 years or more (excluding indeterminate sentences)" ~ ">14 yrs",
      # Custody == "Extended determinate sentence" ~ "",
      # Custody == "Indeterminate sentences" ~ "",
      # Custody == "Recalls" ~ "",
      Custody == "Sentence length not recorded" ~ "Length not recorded",
      Custody == "Non-criminal prisoners(2)" ~ "Non-criminal prisoners"
    ),
    
    Count = as.numeric(Count),
    Date = dmy(Date),
    year = year(Date), month = month(Date),
    #FOR MODELING
    lockdown_happens = ifelse(Date < ymd("2020-03-26") | Date > ymd("2021-06-19"), 0, 1),
    
    #rate of change
    d1_Count = diff(c(0,Count))
         )
```

```{r custody-plot}
###plotting
custody_plot <- Custody_PD_sliced %>%
  group_by(Custody, Gender) %>%
  filter(
    year >= 2015,
    # Age == "Adult",
    Custody == c(
      "Sentenced"
      # "Untried"
    ),
  ) %>% 
  ggplot(
  aes(x = Date, y = d1_Count, group = Gender)) +
  geom_line(aes(group = Gender, colour = Gender),
         ) +
  facet_wrap(
    nrow=6,
    Age + Gender ~ Custody,
    scales = "free_y"
             ) +
  scale_color_manual(values = c(Male = "#19babf",Female = "#ff0048")) +
  labs(x = "Time", y = "Population change") +
  geom_vline(xintercept = ymd("2020-03-26"), linetype = "dotted", size = 0.5) +
  theme(text = element_text(size = 10))
  
# print(custody_plot)

  #in = 100 pixels. Keep commented unless need to export plot
export.width <- 12 
ggsave("img/custody_plot_testing.jpg", plot = custody_plot, width = export.width, height = export.width, units = "in")


  
```

```{r Sky-modeling}

Custody_modeling <- Custody_PD_sliced %>%
  filter(
    year >= 2015,
    Age == "Adult",
    Custody == c(
      "Sentenced"
    )
  )

#Model 1: With interaction
custody_fit_yesInt <- linear_reg() %>%
  set_engine("lm") %>%
  fit(d1_Count ~ Date + lockdown_happens + Date * lockdown_happens, data = Custody_modeling)

#Model 2
custody_fit_noInt <- linear_reg() %>%
  set_engine("lm") %>%
  fit(d1_Count ~ Date + lockdown_happens, data = Custody_modeling)

#Model 3
custody_fit_Simple <- linear_reg() %>%
  set_engine("lm") %>%
  fit(d1_Count ~ lockdown_happens, data = Custody_modeling)

tidy(custody_fit_yesInt)

# Get R-squared and Adjusted R-squared for each model
r_squared_yesInt <- glance(custody_fit_yesInt)$r.squared
r_squared_noInt  <- glance(custody_fit_noInt)$r.squared
r_squared_Simple <- glance(custody_fit_Simple)$r.squared

adj_r_squared_yesInt <- glance(custody_fit_yesInt)$adj.r.squared
adj_r_squared_noInt  <- glance(custody_fit_noInt)$adj.r.squared
adj_r_squared_Simple <- glance(custody_fit_Simple)$adj.r.squared

# Print the results
cat("R-squared for custody_fit_yesInt:", r_squared_yesInt, "\n")
cat("R-squared for custody_fit_noInt:", r_squared_noInt, "\n")
cat("R-squared for custody_fit_Simple:", r_squared_Simple, "\n")
cat("\n")
cat("Adjusted R-squared for custody_fit_yesInt:", adj_r_squared_yesInt, "\n")
cat("Adjusted R-squared for custody_fit_noInt:", adj_r_squared_noInt, "\n")
cat("Adjusted R-squared for custody_fit_Simple:", adj_r_squared_Simple, "\n")


```


# Ethnicity prison data
In this section, prison population is plotted as bar plots with 2 distinct visualisation formats.
<ol>
<li> In the 1st visualisation, colour corresponds to the value of the bars;</li>
<li> In the 2nd visualisation, colour corresponds to the change in population compared to the previous year.</li>
</ol>
```{r ethnicity-plot1}

male_allNat_Ethnic_PD <- Ethnic_PD %>%
  slice(.,34:40) %>% #ignores 'all nationalities' since it is redundant
  select(!2:9)

male_allNat_Ethnic_PD <- as.data.frame(t(male_allNat_Ethnic_PD)) %>%
  row_to_names(row_number = 1) %>%
  rename(
    `Asian_people` = 'Asian/ Asian British(2)',
    Other = 'Other ethnic group(2)',
    `Not_recorded` = 'Ethnic group not recorded',
    `Black_people` = 'Black/ African/ Caribbean/ Black British',
    `Mixed_or_Multiple` = 'Mixed/ Multiple ethnic groups',
    `White_people` = 'White'
  ) %>%
  #converts all columns to numeric
  mutate_all(~ as.numeric(.))

rownames(male_allNat_Ethnic_PD) <- NULL
male_allNat_Ethnic_PD$Year <- c(2009:2023)



### STARTS PLOTTING PRECEDURES

male_allNat_Ethnic_PD %>%
  #adds total
  mutate(All = rowSums(., !Year)) %>%
  #setting up data for the plot
  pivot_longer(!Year, names_to = 'Ethnicity', values_to = 'count') %>%
  group_by(Ethnicity) %>%
  
  #adds percentage
  mutate(group_p = normalise_01(count)) %>%
  
  #plotting
  ggplot(
    aes(
      x = Year,
      y = count,
      fill = group_p
    )
  ) +
  geom_bar(
    aes(
      group = Ethnicity
    ),
    stat = 'identity',
  ) +
  facet_wrap(~Ethnicity,
             scales = 'free_y',
             shrink = FALSE,
             nrow = 2,
             ncol = 4) +
  
  #styling
  theme_light() +
  theme(
    legend.position = 'bottom',
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    x = 'Year',
    y = 'Number of people',
    title = wrapper('Prison population in England and Wales, grouped by ethnicity, from year 2009 to 2023', 60),
    fill = wrapper('normalised indicator of max and min within Ethnicity group',40)
  ) +
  # scale_fill_gradient(low='red', high='forestgreen', breaks = c(0,1))
  scale_fill_gradient(low='violetred3', high='chartreuse3',
                         breaks = c(0,0.5,1),
                       labels = c('min','','max')
                      )
```

```{r ethnicity-plot2}
  #plotting with different viz, namely difference between years
  
  male_allNat_Ethnic_PD %>%
  mutate(All = rowSums(., !Year)) %>%
  pivot_longer(!Year, names_to = 'Ethnicity', values_to = 'count') %>%
  group_by(Ethnicity) %>%
  
  mutate(group_p = c(0,normalise_01(diff(count,lag = 1)) ) ) %>%
  
  ### plot
  ggplot(
    aes(
      x = Year,
      y = count,
      fill = group_p
    )
  ) +
  geom_bar(
    aes(
      group = Ethnicity
    ),
    stat = 'identity',
  ) +
  facet_wrap(~Ethnicity,
             scales = 'free_y',
             shrink = FALSE,
             nrow = 2,
             ncol = 4) +
  annotate(
    "rect",
    xmin=2020-0.5, xmax=2022+0.5,
    ymin=-Inf, ymax=Inf,
    alpha = 0.2, fill = "black", color = "black"
  ) +
  
  ### styling
  theme_light() +
  theme(
    legend.position = 'bottom',
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    x = 'Year',
    y = 'Number of people',
    title = wrapper('Prison population in England and Wales, grouped by ethnicity, from year 2009 to 2023', 60),
    fill = wrapper('normalised indicator of difference from previous year within Ethnicity group',40)
  ) +
  scale_fill_gradient2(low='violetred3', mid='grey', high='chartreuse3',
                       midpoint = 0,
                         breaks = c(-1,0,1),
                       labels = c('decrease','no\n change','increase')
                      )
```

# Total prison data  
In this section, prison population is plotted as line graphs  
<ol>
<li> The first visualisation shows total prison population by type of custody </li>
<li> The second visualisations shows the difference from the previous data point (note that after 2009 every quarter is used rather than every year)</li>
<li> The third visualisation shows the rate of change weekly from the previous data point</li>
</ol>


```{r basic A1_7 plotting,fig.width=10, fig.height=5}

Annual_Data_A1_7 <- as.data.frame(A1_7)

#selects wanted data and the dates
Annual_Data_A1_7 <- Annual_Data_A1_7 %>%
  slice(1,2,13,24,35)

#cleans names
Annual_Data_A1_7[5,1] <- "Non-criminal prisoners"
Annual_Data_A1_7[2,1] <- "Total"

#makes the column names the Date
colnames(Annual_Data_A1_7)[-1] <- Annual_Data_A1_7[1,-1]

colnames(Annual_Data_A1_7)[1] <- "Group"

#pivots the data frame creating 3 columns being "Group","Date","Arrests"
Annual_Data_A1_7 <- Annual_Data_A1_7 %>%
  slice(-1) %>%
  pivot_longer(!Group, names_to = "Date", values_to = "Arrests") %>%
  mutate(Date = as.Date(Date, format = "%d-%m月-%y"),
         Arrests = as.numeric(Arrests)) 

Annual_Data_A1_7 <- na.omit(Annual_Data_A1_7)

#basic data plot by Group
ggplot(data = (Annual_Data_A1_7),
       mapping = aes( 
         x = Date,
         y = Arrests,
         group = Group,
         colour = Group)) + 
  geom_line(size = 1) + 
  facet_wrap(~Group, scales = "free") 
```

```{r A1_7 change plotting,fig.width=10, fig.height=5}
#Calculates the % change from previous value (note first value is set to NA)
diff_Annual_Data_A1_7 <- Annual_Data_A1_7 %>%
  filter(year(Date) > 2009) %>%
  arrange(Group, Date) %>%
  group_by(Group) %>%
  mutate(
    Arrests_diff = c(NA, diff(Arrests)),
    Arrests_pct_change = (Arrests_diff / lag(Arrests)) * 100
    ) %>%
  ungroup
Total_group_data <- diff_Annual_Data_A1_7 %>%
  filter(Group == "Total")
#plots difference between each data point by group
ggplot(data = na.omit(Total_group_data),
       mapping = aes( 
         x = Date,
         y = Arrests_pct_change,
         group = Group,
         colour = Group)) + 
  geom_line(size = 1) + 
  annotate(
    "rect",
    xmin= as.Date("2012-01-01"), xmax=as.Date("2014-06-01"),
    ymin=-Inf, ymax=Inf,
    alpha = 0.2, fill = "black", color = "black"
  ) +
  annotate(
    "rect",
    xmin= as.Date("2017-03-01"), xmax=as.Date("2020-01-01"),
    ymin=-Inf, ymax=Inf,
    alpha = 0.2, fill = "black", color = "black"
  ) +
  facet_wrap(~Group, scales = "free")

 
```

```{r Gradient of A1_7,fig.width=10, fig.height=5}
#quarter is measured in days / 91.3125 (Ave days in a quarter)
grad_Annual_Data_A1_7 <- diff_Annual_Data_A1_7 %>%
  filter(year(Date) > 2009) %>%
  arrange(Group, Date) %>%
  group_by(Group) %>%
  mutate(
    Date_diff = c(NA, as.numeric(difftime(Date[-1], Date[-length(Date)], units = "days")/ 91.3125)),
    Gradient = Arrests_diff / Date_diff
    ) %>%
  ungroup()
Total_group_data <- grad_Annual_Data_A1_7 %>%
  filter(Group == "Total")

#plots gradient by group
ggplot(data = na.omit(Total_group_data),
       mapping = aes( 
         x = Date,
         y = Gradient,
         group = Group,
         colour = Group)) + 
  geom_line(size = 1) + 
  annotate(
    "rect",
    xmin= as.Date("2012-01-01"), xmax=as.Date("2014-06-01"),
    ymin=-Inf, ymax=Inf,
    alpha = 0.2, fill = "black", color = "black"
  ) +
  annotate(
    "rect",
    xmin= as.Date("2017-03-01"), xmax=as.Date("2020-01-01"),
    ymin=-Inf, ymax=Inf,
    alpha = 0.2, fill = "black", color = "black"
  ) +
  facet_wrap(~Group, scales = "free") 
```

```{r 2012-17 type of crime and sentence type,fig.width=10, fig.height=5}
Annual_Data_A1_5ii <- as.data.frame(TypeOfCrime)
Annual_Data_A1_5i <- as.data.frame(A1_5i)
Annual_Data_A1_1 <- as.data.frame(Custody_PD)

#selects wanted data and the dates
Annual_Data_A1_5ii <- Annual_Data_A1_5ii %>%
  slice(1, #date
    13, #rape
    16,#robbery
    17, #burglary
    18, #theft and handling
    22, #fraud and forgery
    25) #drug offences
Annual_Data_A1_5i <- Annual_Data_A1_5i %>%
  slice(1, #date
        34, #rape
        61,#robbery
        62, #theft offences
        89, #drug offences
        128)
Annual_Data_A1_1 <- Annual_Data_A1_1 %>%
  slice(1,3,4,5,7,21)

#makes the column names the Date
colnames(Annual_Data_A1_5ii)[-1] <- Annual_Data_A1_5ii[1,-1]
colnames(Annual_Data_A1_5ii)[1] <- "Group"
colnames(Annual_Data_A1_5i)[-1] <- Annual_Data_A1_5i[1,-1]
colnames(Annual_Data_A1_5i)[1] <- "Group"
colnames(Annual_Data_A1_1)[-1] <- Annual_Data_A1_1[1,-1]
colnames(Annual_Data_A1_1)[1] <- "Group"

#pivots the data frame creating 3 columns being "Group","Date","Arrests"
Annual_Data_A1_5ii <- Annual_Data_A1_5ii %>%
  slice(-1) %>%
  pivot_longer(!Group, names_to = "Date", values_to = "Arrests") %>%
  mutate(Date = as.Date(Date, format = "%d-%m月-%y"),
         Arrests = as.numeric(Arrests))
Annual_Data_A1_5i <- Annual_Data_A1_5i %>%
  slice(-1) %>%
  pivot_longer(!Group, names_to = "Date", values_to = "Arrests") %>%
  mutate(Date = as.Date(Date, format = "%d-%m月-%y"),
         Arrests = as.numeric(Arrests)) 
Annual_Data_A1_1 <- Annual_Data_A1_1 %>%
  slice(-1) %>%
  pivot_longer(!Group, names_to = "Date", values_to = "Arrests") %>%
  mutate(Date = as.Date(Date, format = "%d-%m月-%y"),
         Arrests = as.numeric(Arrests)) 

Annual_Data_A1_5ii <- na.omit(Annual_Data_A1_5ii)
Annual_Data_A1_5i <- na.omit(Annual_Data_A1_5i)
Annual_Data_A1_1 <- na.omit(Annual_Data_A1_1)

#merge datas
Annual_Data_A1_5ii <- Annual_Data_A1_5ii %>%
  mutate(Group = if_else(
    Group %in% c("Theft and handling","Burglary"),"Theft Offences", Group)) %>%
  group_by(Group, Date) %>%
  summarise(Arrests = sum(Arrests, na.rm = TRUE), .groups = "drop") %>%
  mutate(Group = if_else(Group == "Fraud and forgery","Fraud Offences", Group))

Annual_Date_A1_5 <- Annual_Data_A1_5ii %>%
  full_join(Annual_Data_A1_5i)

#filter years
Annual_Date_A1_5 <- Annual_Date_A1_5 %>%
  filter(Date > as.Date("2010-05-11"),
         Date < as.Date("2019-07-24"))
Annual_Date_A1_1 <- Annual_Data_A1_1 %>%
  filter(Date > as.Date("2010-05-11"),
         Date < as.Date("2019-07-24"))

#basic data plot by Group
ggplot(data = (Annual_Date_A1_5),
       mapping = aes( 
         x = Date,
         y = Arrests,
         group = Group,
         colour = Group)) + 
  geom_line(size = 1) + 
  labs(title = "Offence Type",
       x = "Date",
       y = "Population") +
  facet_wrap(~Group, scales = "free") 

ggplot(data = (Annual_Date_A1_1),
       mapping = aes( 
         x = Date,
         y = Arrests,
         group = Group,
         colour = Group)) + 
  geom_line(size = 1) + 
labs(title = "Remand",
       x = "Date",
       y = "Population") +
  facet_wrap(~Group, scales = "free") 
```

```{r analysis of penal reforms}
#this has repeat code but this can therefore be used as a template
#load data
PR_A1_1 <- as.data.frame(Custody_PD)
PR_A1_7 <- as.data.frame(A1_7)

#selects wanted data and the dates
PR_A1_1 <- PR_A1_1 %>%
  slice(1,2,3,6,25,28,96,190)
PR_A1_7 <- PR_A1_7 %>%
  slice(1,3,4,5,6,7,8,9,10)

#makes the column names the Date and changes col name (col 1 changes based on data)
colnames(PR_A1_1)[-1] <- PR_A1_1[1,-1]
colnames(PR_A1_1)[1] <- "Custody_type"
colnames(PR_A1_7)[-1] <- PR_A1_7[1,-1]
colnames(PR_A1_7)[1] <- "Age"

#pivots the data frame creating 3 columns
PR_A1_1 <- PR_A1_1 %>%
  slice(-1) %>%
  pivot_longer(!Custody_type, names_to = "Date", values_to = "Prison_pop") %>%
  mutate(Date = as.Date(Date, format = "%d-%m月-%y"),
         Prison_pop = as.numeric(Prison_pop)) %>%
  distinct(Date, Custody_type, .keep_all = TRUE)
PR_A1_7 <- PR_A1_7 %>%
  slice(-1) %>%
  pivot_longer(!Age, names_to = "Date", values_to = "Prison_pop") %>%
  mutate(Date = as.Date(Date, format = "%d-%m月-%y"),
         Prison_pop = as.numeric(Prison_pop)) %>%
   distinct(Date, Age, .keep_all = TRUE)
PR_A1_1 <- na.omit(PR_A1_1)
PR_A1_7 <- na.omit(PR_A1_7)

#to make code neater i added gender and total into PR_A1_1 and filtered time here
#filter to selected time indicator (this case is monthly)

PR_A1_7 <- PR_A1_7 %>%
  filter(Date >= as.Date("2010-06-01"),
         Date <= as.Date("2020-06-01"))
PR_A1_1 <- PR_A1_1 %>%
  filter(Date >= as.Date("2010-06-01"),
         Date <= as.Date("2020-06-01"))

#add a new data frame for total
PR_total <- PR_A1_1 %>%
  filter(Custody_type == "Males and Females")
colnames(PR_total)[3] <- "Total_pop"
PR_total <- PR_total %>%
  select(!1)

#new data frame for males then females
PR_male <- PR_A1_1 %>%
    filter(Custody_type == "Males")
colnames(PR_male)[3] <- "Male_pop"
PR_male <- PR_male %>%
  select(!1)
PR_female <- PR_A1_1 %>%
    filter(Custody_type == "Females")
colnames(PR_female)[3] <- "Female_pop"
PR_female <- PR_female %>%
  select(!1)

#custody type data frames
PR_Remand <- PR_A1_1 %>%
  filter(Custody_type == "Remand")
colnames(PR_Remand)[3] <- "Remand_pop"
PR_Remand <- PR_Remand %>%
  select(!1)
PR_Sentenced <- PR_A1_1 %>%
  filter(Custody_type == "Sentenced")
colnames(PR_Sentenced)[3] <- "Sentenced_pop"
PR_Sentenced <- PR_Sentenced %>%
  select(!1)

#age data frames
PR_age <- PR_A1_7 %>%
  mutate(Age = case_when(
     Age %in% c("15-17","18-20", "21-24") ~ "Under 25",
      Age %in% c("25-29", "30-39") ~ "25-40",
      Age %in% c("40-49","50-59") ~ "40-60",
      Age %in% c("60 and over") ~ "60+",
      TRUE ~ "Unknown")) %>%
  group_by(Age,Date) %>%
  summarise(Prison_pop = sum(Prison_pop), .groups = "drop")

PR_u25 <- PR_age %>%
  filter(Age == "Under 25")

colnames(PR_u25)[3] <- "Under_25_pop"
PR_u25 <- PR_u25 %>%
  select(!1)

PR_25_40 <- PR_age %>%
  filter(Age == "25-40")
colnames(PR_25_40)[3] <- "at_25_40_pop"

PR_25_40 <- PR_25_40 %>%
  select(!1)

PR_40_60 <- PR_age %>%
  filter(Age == "40-60")
colnames(PR_40_60)[3] <- "at_40_60_pop"

PR_40_60 <- PR_40_60 %>%
  select(!1)

PR_a60 <- PR_age %>%
  filter(Age == "60+")

colnames(PR_a60)[3] <- "Above_60_pop"
PR_a60 <- PR_a60 %>%
  select(!1)

#list of data frames
dfs <- list(PR_female, PR_male, PR_Remand, PR_Sentenced, PR_u25, PR_25_40, PR_40_60, PR_a60, PR_total)

#use reduce to join all data frames in the list by "Date"
PR_quarterly <- reduce(dfs, left_join, by = "Date")
PR_quarterly <- na.omit(PR_quarterly)

dfs_2 <- list(PR_female, PR_male, PR_Remand, PR_Sentenced, PR_total)
PR_monthly <- reduce(dfs_2, left_join, by = "Date")

#event dates
event_dates <- as.Date(c("2012-05-01", "2013-04-25", "2014-03-13", "2015-02-12"))
event_quarters <- quarter(event_dates, with_year = TRUE)

#checking if in same quarter
PR_quarterly <- PR_quarterly %>%
  mutate(event_occurred = if_else(quarter(Date, with_year = TRUE) %in% event_quarters, 1, 0))

PR_monthly <- PR_monthly %>%
  rowwise() %>%
  mutate(event_occurred = any(Date > event_dates & Date <= event_dates + months(8))) %>%
  ungroup()

#t test (messures mean before and after events)
PR_t_test <- t.test(Total_pop ~ event_occurred, data = PR_monthly) %>%
  tidy()
print(PR_t_test)

#t test is mostly not visualisable so this can kinda be used as a visual representation
ggplot(PR_monthly, aes(x = as.factor(event_occurred), y = Total_pop, fill = as.factor(event_occurred))) +
  geom_boxplot() +
  labs(title = "Boxplot of Total Population by Event Occurrence",
       x = "Event Occurred",
       y = "Total Population") +
  theme_minimal()

#convert Date to numeric for regression modelling
PR_quarterly$Date_num <- as.numeric(as.Date(PR_quarterly$Date))

#linear regression model with Date and Event indicator for quarters
PR_model <- linear_reg() %>%
  set_engine("lm") %>%
  fit(Total_pop ~ Date_num + event_occurred, data = PR_quarterly) 
model_summary <- summary(PR_model$fit)
model_summary
# Print R-squared and Adjusted R-squared
r_squared <- model_summary$r.squared
adj_r_squared <- model_summary$adj.r.squared
r_squared
adj_r_squared 

#time series analysis (can be used to see the difference in predicted vs actual)
PR_monthly_to_2012 <- PR_monthly %>%
  filter(Date < as.Date("2012-05-01"))

ts_data <- ts(PR_monthly_to_2012$Total_pop, start = c(2010, 7), frequency = 12) #number of months in a year
fit_arima <- auto.arima(ts_data)
forecast_result <- forecast(fit_arima, h = 48) 

#clean ts data
forecast_df <- rownames_to_column(as.data.frame(forecast_result), var = "Date") %>%
  mutate(Date = as.yearmon(Date, format = "%b %Y"),
         Date = as.Date(Date, format = "%b %y", frac = 1)) #end of month
ts_vs_real <- inner_join(forecast_df, PR_monthly , by = "Date")

ts_vs_real %>%
  filter(Date <= as.Date("2016-09-01")) %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(y = `Point Forecast`, color = "Point Forecast")) +
  geom_line(aes(y = `Total_pop`, color = "Total_pop")) +
  scale_color_manual(values = c("Point Forecast" = "blue", "Total_pop" = "red")) +
  labs(title = "Forecast vs Real",
       x = "Date", 
       y = "Population")
```


```{r,fig.width=10, fig.height=5}
Lockdown_data <- Annual_Data_A1_7
Lockdown_data$prison_population <- Lockdown_data$Arrests

# Ensure the Date column is in Date format
Lockdown_data$Date <- as.Date(Lockdown_data$Date)

# Define lockdown periods 
Lockdown_data$lockdown <- ifelse(
  (Lockdown_data$Date >= "2020-03-23" & Lockdown_data$Date <= "2020-07-04") |  # First lockdown
  (Lockdown_data$Date >= "2020-11-05" & Lockdown_data$Date <= "2020-12-02") |  # Second lockdown
  (Lockdown_data$Date >= "2021-01-06" & Lockdown_data$Date <= "2021-03-29"),   # Third lockdown
  1,  # Assign 1 during lockdown
  0   # Assign 0 otherwise
)

library(tidymodels)            

# Fit the linear model using tidymodels syntax
model <- linear_reg() %>%
  set_engine("lm") %>%
  fit(prison_population ~ lockdown, 
      data = Lockdown_data)

# View the model coefficients
tidy(model)



```

```{r Type of Crime,fig.width=10, fig.height=5}
#hanna and vivi's
Type_of_Crime_Graph <- as.data.frame(TypeOfCrime)

#selects wanted data and the dates
Type_of_Crime_Graph <- Type_of_Crime_Graph %>%
  slice(1,2,3,13,16,17,22,25,31,34)

#cleans names
#Annual_Data_A1_7[5,1] <- "Non-criminal prisoners"
#Annual_Data_A1_7[2,1] <- "Total"

#makes the column names the Date
colnames(Type_of_Crime_Graph)[-1] <- Type_of_Crime_Graph[1,-1]

colnames(Type_of_Crime_Graph)[1] <- "Group"

#pivots the data frame creating 3 columns being "Group","Date","Arrests"
Type_of_Crime_Graph <- Type_of_Crime_Graph %>%
  slice(-1) %>%
  pivot_longer(!Group, names_to = "Date", values_to = "Arrests") %>%
  mutate(Date = as.Date(Date, format = "%d-%m月-%y"),
         Arrests = as.numeric(Arrests)) 

Type_of_Crime_Graph <- na.omit(Type_of_Crime_Graph)

#basic data plot by Group
ggplot(data = (Type_of_Crime_Graph),
       mapping = aes( 
         x = Date,
         y = Arrests,
         group = Group,
         colour = Group)) + 
  geom_line(size = 1) + 
  facet_wrap(~Group, scales = "free")  

PM_A1_1 <- as.data.frame(Custody_PD)
PM_A1_7 <- as.data.frame(A1_7)

#selects wanted data and the dates
PM_A1_1 <- PM_A1_1 %>%
  slice(1,2,3,6,25,28,96,190)
PM_A1_7 <- PM_A1_7 %>%
  slice(1,3,4,5,6,7,8,9,10)

#makes the column names the Date and changes col name (col 1 changes based on data)
colnames(PM_A1_1)[-1] <- PM_A1_1[1,-1]
colnames(PM_A1_1)[1] <- "Custody_type"
colnames(PM_A1_7)[-1] <- PM_A1_7[1,-1]
colnames(PM_A1_7)[1] <- "Age"

#pivots the data frame creating 3 columns
PM_A1_1 <- PM_A1_1 %>%
  slice(-1) %>%
  pivot_longer(!Custody_type, names_to = "Date", values_to = "Prison_pop") %>%
  mutate(Date = as.Date(Date, format = "%d-%m月-%y"),
         Prison_pop = as.numeric(Prison_pop)) %>%
  distinct(Date, Custody_type, .keep_all = TRUE)
PM_A1_7 <- PM_A1_7 %>%
  slice(-1) %>%
  pivot_longer(!Age, names_to = "Date", values_to = "Prison_pop") %>%
  mutate(Date = as.Date(Date, format = "%d-%m月-%y"),
         Prison_pop = as.numeric(Prison_pop)) %>%
   distinct(Date, Age, .keep_all = TRUE)

PM_A1_1 <- na.omit(PM_A1_1)
PM_A1_7 <- na.omit(PM_A1_7)

#to make code neater i added gender and total into PR_A1_1 and filtered time here
#filter to selected time indicator (this case is monthly)

PM_A1_7 <- PM_A1_7 %>%
  filter(Date >= as.Date("2010-05-11"),
         Date <= as.Date("2019-07-24"))
PM_A1_1 <- PM_A1_1 %>%
  filter(Date >= as.Date("2010-05-11"),
         Date <= as.Date("2019-07-24"))

#add a new data frame for total
PM_total <- PM_A1_1 %>%
  filter(Custody_type == "Males and Females")
colnames(PM_total)[3] <- "Total_pop"
PM_total <- PM_total %>%
  select(!1)

#new data frame for males then females
PM_male <- PM_A1_1 %>%
    filter(Custody_type == "Males")
colnames(PM_male)[3] <- "Male_pop"
PM_male <- PM_male %>%
  select(!1)
PM_female <- PM_A1_1 %>%
    filter(Custody_type == "Females")
colnames(PM_female)[3] <- "Female_pop"
PM_female <- PM_female %>%
  select(!1)

#custody type data frames
PM_Remand <- PM_A1_1 %>%
  filter(Custody_type == "Remand")
colnames(PM_Remand)[3] <- "Remand_pop"
PM_Remand <- PM_Remand %>%
  select(!1)
PM_Sentenced <- PM_A1_1 %>%
  filter(Custody_type == "Sentenced")
colnames(PM_Sentenced)[3] <- "Sentenced_pop"
PM_Sentenced <- PM_Sentenced %>%
  select(!1)

#age data frames
PM_age <- PM_A1_7 %>%
  mutate(Age = case_when(
     Age %in% c("15-17","18-20", "21-24") ~ "Under 25",
      Age %in% c("25-29", "30-39") ~ "25-40",
      Age %in% c("40-49","50-59") ~ "40-60",
      Age %in% c("60 and over") ~ "60+",
      TRUE ~ "Unknown")) %>%
  group_by(Age,Date) %>%
  summarise(Prison_pop = sum(Prison_pop), .groups = "drop")

PM_u25 <- PM_age %>%
  filter(Age == "Under 25")

colnames(PM_u25)[3] <- "Under_25_pop"
PM_u25 <- PM_u25 %>%
  select(!1)

PM_25_40 <- PM_age %>%
  filter(Age == "25-40")
colnames(PM_25_40)[3] <- "at_25_40_pop"

PM_25_40 <- PM_25_40 %>%
  select(!1)

PM_40_60 <- PM_age %>%
  filter(Age == "40-60")
colnames(PM_40_60)[3] <- "at_40_60_pop"

PM_40_60 <- PM_40_60 %>%
  select(!1)

PM_a60 <- PM_age %>%
  filter(Age == "60+")

colnames(PM_a60)[3] <- "Above_60_pop"
PM_a60 <- PM_a60 %>%
  select(!1)

#list of data frames
dfs <- list(PM_female, PM_male, PM_Remand, PM_Sentenced, PM_u25, PM_25_40, PM_40_60, PM_a60, PM_total)

#use reduce to join all data frames in the list by "Date"
PM_quarterly <- reduce(dfs, left_join, by = "Date")
PM_quarterly <- na.omit(PM_quarterly)

dfs_2 <- list(PM_female, PM_male, PM_Remand, PM_Sentenced, PM_total)
PM_monthly <- reduce(dfs, left_join, by = "Date")

#event date
event_date <- as.Date("2016-07-13")

#checking if its after event_date
PM_quarterly <- PM_quarterly %>%
  mutate(event_occurred = ifelse(Date < event_date, 0, 1))

PM_monthly <- PM_monthly %>%
  mutate(event_occurred = ifelse(Date < event_date, 0, 1))

#t test (measures mean before and after events)
PM_t_test <- t.test(Total_pop ~ event_occurred, data = PM_monthly) %>%
  tidy()
print(PM_t_test)

#t test is mostly not visualisable so this can kinda be used as a visual representation
ggplot(PM_monthly, aes(x = as.factor(event_occurred), y = Total_pop, fill = as.factor(event_occurred))) +
  geom_boxplot() +
  labs(title = "Boxplot of Total Population by Event Occurrence",
       x = "Event Occurred",
       y = "Total Population") +
  theme_minimal()

#convert Date to numeric for regression modelling
PM_quarterly$Date_num <- as.numeric(as.Date(PM_quarterly$Date))

#linear regression model with Date and Event indicator for quarters
PM_model <- linear_reg() %>%
  set_engine("lm") %>%
  fit(Total_pop ~ Date_num + event_occurred, data = PM_quarterly) 
model_summary <- summary(PM_model$fit)
model_summary
# Print R-squared and Adjusted R-squared
r_squared <- model_summary$r.squared
adj_r_squared <- model_summary$adj.r.squared
r_squared
adj_r_squared 

#time series analysis (can be used to see the difference in predicted vs actual)
#number of quarters in a year
PM_monthly_to_2012 <- PM_monthly %>%
  filter(Date >= as.Date("2010-04-01") & Date <= as.Date("2016-07-13"))
ts_data <- ts(PM_monthly_to_2012$Total_pop,  start = c(2010, 2), frequency = 12)  #number of quarters in a year
fit_arima <- auto.arima(ts_data)
forecast_result <- forecast(fit_arima, h = 36) 

#forecasting the next 12 periods
print(forecast_result)

#clean ts data
forecast_df <- rownames_to_column(as.data.frame(forecast_result), var = "Date") %>%
  mutate(Date = as.yearmon(Date, format = "%b %Y"),
         Date = as.Date(Date, format = "%b %y", frac = 1)) #end of month
ts_vs_real <- inner_join(forecast_df, PM_monthly , by = "Date")

# Ensure the ts_vs_real data frame has the correct structure
ts_vs_real %>%
  filter(Date <= as.Date("2019-07-24")) %>%
  ggplot(aes(x = Date)) +
  geom_line(aes(
    y = `Point Forecast`, color = "Point Forecast")) +
  geom_line(aes(y = `Hi 95`, color = "Hi 95")) +
  geom_line(aes(y = `Lo 95`, color = "Lo 95")) +
  geom_line(aes(y = Total_pop, color = "Total_pop")) +
  scale_color_manual(values = c("Point Forecast" = "blue", 
                                "Lo 95" = "orange", 
                                "Hi 95" = "orange", 
                                "Total_pop" = "red")) +
  labs(title = "Forecast vs Real",
       x = "Date", 
       y = "Population") 

```