---
title: "Correlation Between Significant Events and Prison Population"
author: "by Gomakashi: Marcel, Sky, Hanna, Vivi, Siddhi & Maggie"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
library(ggplot2)

#PACKAGES YOU MIGHT NEED TO INSTALL
library(readODS)
library(janitor) #for changing row names
library(RColorBrewer)

```


```{r load-data, include=FALSE}
Ethnic_PD <- read_ods('data/Population_30June2023_Annual.ods', sheet = 14)

#wrapper function for generating title of plots
wrapper <- function(x, ...) 
{
  paste(strwrap(x, ...), collapse = '\n')
}

#normalises data to range from 0 to 1
normalise_01 <- function(x)
{
  if (min(x) < 0 )
  {
    x/max(abs(x))
  } else {
    (x-min(x))/(max(x)-min(x))
  }
}


```


## Research Question: Correlation between significant events and prison population of England and Wales

We have picked some major worldwide events to see whether if there is a noticeable correlation between them and changes in prison population.

## Data

We explored the 'Prison population by type of custody, age group and sex' section of the data.

The data set we looked at contained many custody groups like remand(individuals that are held in prison until their trial/hearing), intermediate sentence(individuals that have to stay in custody until they have been approved to be safe for the public), fine defaulter(individuals that are in prison due to refusing to pay a fine) and recalls(individuals that were released on parole/license but broke the rules of their probation and are back in prison). 

Note to future editor:
Elaborate in this section on:
- the time resolution (how much time is spaced between consecutive data), how it went from once a year to once a month in 2011, etc.)
- explaination of other data set used

## Findings
### Prime Minister
For the Prime Minister section, we analyzed the full data set across all years to compare crime trends before and during each Prime Minister's term. However, limitations included difficulty in directly linking changes to specific policies, leading to some assumptions. To address this, we examined different crime types and related policies.

We then looked into key issues and possible related government responses for both Lord Cameron and Theresa May.

#### Lord Cameron(2010-2016):

##### 1.Drug Offences

Trend: 
Significant decline after 2010

Key Issues:

- Shift from punitive measures to rehabilitation and tackling high-level trafficking.

Government Response:

- Expanded funding for drug treatment and harm reduction initiatives.

##### 2.Fraud Offences

Trend: 
Declined, reflecting a transition to cyber fraud.

Key Issues:

- Limited resources allocated for investigating complex fraud and increasing online scams

Government Response:

- Strengthened cybercrime units and encouraged private sector involvement in fraud prevention. 

##### 3.Rape

Trend: 
Reports and imprisonments steadily increased.

Key Issues:

- Greater awareness and public demand for action

Government Response:

- Funded support services and introduced training for handling rape cases.


##### 4.Robbery

Trend: 
Peaked early, then declined significantly

Key Issues:

- Economic recession initially drove rates up, while urban safety measures later reduced offenses.

Government Response:

- Promoted urban safety initiatives like CCTV and improved penalties for violent robbery.


##### 5.Theft Offences 

Trend: 
Declined after an intial rise.

Key Issues:

- Technological advancements in security deterred theft.

Government Response:

- Supported public safety campaigns and neighbourhood watch programs.


#### Theresa May(2016-2019):

##### 1.Drug Offences

Trend:
Fluctuated, with a rise linked to county lines operations.

Key Issues:

- Vulnerable individuals exploited by organized crime.

Government Response:

- Strengthened the National Crime Agency (NCA) and introduced targeted interventions.


##### 2.Fraud Offenses

Trend: 
Continue declined due to a shift towards cyber fraud. 

Key Issues:

- Limited resources for complex fraud investigations.

Government Response:

- Promoted private sector collaboration and expanded cybercrime units.


##### 3.Rape

Trend: 
Reports rose sharply, reflecting increased awareness

Key Issues:

- Persistent gaps in convictions despite higher reporting.

Government Response:

- Improved victim support and police training for sensitive cases.

##### 4.Robbery

Trend: 
Steep decline due to improved surveillance and urban safety measures.

Key Issues:

- Economic challenges and reduced police presence.

Government Response:

- Strengthened penalties and promoted urbvan safety programs.


##### 5.Theft Offences

Trend: 
Consistently declined due to imporved security measures.

Key Issues:

- Shift from physical theft to online fraud

Government Response:

- Supported neighborhood watch programs and public awareness campaigns. 









The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.




## References

Ministry of Justice, Offender Management Statistics Quarterly publications, "Prison population: 30 June 2023", Retrieved November 8, 2024, Available at: https://www.gov.uk/government/collections/offender-management-statistics-quarterly

