---
title: "Correlation Between Significant Events and Prison Population"
author: "by Gomakashi: Marcel, Sky, Hanna, Vivi, Siddhi & Maggie"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
library(ggplot2)

#PACKAGES YOU MIGHT NEED TO INSTALL
library(readODS)
library(janitor) #for changing row names
library(RColorBrewer)
library(broom)

```


```{r load-data, include=FALSE}

```


## Research Question: Correlation between significant events and prison population of England and Wales

We have picked some major worldwide events to see whether if there is a noticeable correlation between them and changes in prison population.

## Findings
### Prime Minister
For the Prime Minister section, we analyzed the full data set across all years to compare crime trends before and during each Prime Minister's term. However, limitations included difficulty in directly linking changes to specific policies, leading to some assumptions. To address this, we examined different crime types and related policies.

We then looked into key issues and possible related government responses for both Lord Cameron and Theresa May.

#### Lord Cameron(2010-2016):

##### 1.Drug Offences

Trend: 
Significant decline after 2010

Key Issues:

- Shift from punitive measures to rehabilitation and tackling high-level trafficking.

Government Response:

- Expanded funding for drug treatment and harm reduction initiatives.

##### 2.Fraud Offences

Trend: 
Declined, reflecting a transition to cyber fraud.

Key Issues:

- Limited resources allocated for investigating complex fraud and increasing online scams

Government Response:

- Strengthened cybercrime units and encouraged private sector involvement in fraud prevention. 

##### 3.Rape

Trend: 
Reports and imprisonments steadily increased.

Key Issues:

- Greater awareness and public demand for action

Government Response:

- Funded support services and introduced training for handling rape cases.


##### 4.Robbery

Trend: 
Peaked early, then declined significantly

Key Issues:

- Economic recession initially drove rates up, while urban safety measures later reduced offenses.

Government Response:

- Promoted urban safety initiatives like CCTV and improved penalties for violent robbery.


##### 5.Theft Offences 

Trend: 
Declined after an intial rise.

Key Issues:

- Technological advancements in security deterred theft.

Government Response:

- Supported public safety campaigns and neighbourhood watch programs.


#### Theresa May(2016-2019):

##### 1.Drug Offences

Trend:
Fluctuated, with a rise linked to county lines operations.

Key Issues:

- Vulnerable individuals exploited by organized crime.

Government Response:

- Strengthened the National Crime Agency (NCA) and introduced targeted interventions.


##### 2.Fraud Offenses

Trend: 
Continue declined due to a shift towards cyber fraud. 

Key Issues:

- Limited resources for complex fraud investigations.

Government Response:

- Promoted private sector collaboration and expanded cybercrime units.


##### 3.Rape

Trend: 
Reports rose sharply, reflecting increased awareness

Key Issues:

- Persistent gaps in convictions despite higher reporting.

Government Response:

- Improved victim support and police training for sensitive cases.

##### 4.Robbery

Trend: 
Steep decline due to improved surveillance and urban safety measures.

Key Issues:

- Economic challenges and reduced police presence.

Government Response:

- Strengthened penalties and promoted urbvan safety programs.


##### 5.Theft Offences

Trend: 
Consistently declined due to imporved security measures.

Key Issues:

- Shift from physical theft to online fraud

Government Response:

- Supported neighborhood watch programs and public awareness campaigns. 


## A look at the effect of COVID-19 on prison population of different custody types

### How lockdown affected the prisoner population
We analysed the rate of change in the prisoner population (d1_Count) over time, using a subset of our data that is only the adult prisoners in sentenced custody, since it is the most populous category in our data and roughly reflects all other age subsets.

#### Methods
We created an indicator variable `lockdown_happens` which is 1 during the first lockdown period (2020 March 26 to July 4), and 0 at any other given time. This will be used as the explanatory variable for our model.

We used linear regression for our models, the response variable being the rate of change of prisoner population `d1_Count`, and the explanatory variables being `lockdown_happens` and `Date`. `Date` is included to test whether there was a trend with respect to time in `d1_Count`.

1. **Model 1 (custody_fit_yesInt)**: This model included the variables `Date`, `lockdown_happens`, and the interaction between these two variables (`Date * lockdown_happens`). This model assesses how the lockdown period affected the population change in sentenced custody over time while accounting for any general time trend.

2. **Model 2 (custody_fit_noInt)**: This model is did not include interaction term and only considers the effects of `Date` and `lockdown_happens`.

3. **Model 3 (custody_fit_Simple)**: The simplest model of the three, which only uses `lockdown_happens` as a predictor. Since `lockdown_happens` takes the value of either 0 or 1, the modelled line of fit would be horizontal lines. This was considered because from preliminary assessment of the data through graphs, we saw that although `d1_Count` is unstable, it oscillates roughly around 0 throughout time, and in some cases like Extended determinate sentences, there appeared to be a decreasing trend with respect to time.

#### Results

**Model 1 (custody_fit_yesInt)** The estimates for Date and the interaction term are negligible and statistically insignificant (p-values > 0.7). The lockdown_happens estimate is large (17006) but not significant (p = 0.738). It has the lowest adjusted R-squared (0.1663), suggesting minimal explanatory power for the interaction effect. 

**Model 2 (custody_fit_noInt)** Without interaction, the lockdown_happens variable retains a strong negative estimate (-632, p < 0.001), while Date’s estimate (0.00795) is not significant (p = 0.702). It has a slightly higher adjusted R-squared (0.1699), indicating that the interaction term does not add significant value.

**Model 3 (custody_fit_Simple)** With only lockdown_happens, it shows a strong negative estimate for lockdown_happens (-629, p < 0.001). It has the highest adjusted R-squared (0.1734), showing that lockdown_happens is the key factor of variance in d1_Count without additional predictors or interactions.

#### Interpretation

Using Occam's Razor, the simplest model that still provides meaningful explanatory power should be used. Since the r-squared of Model 3 is the highest and simplest, this makes it the most interpretable for directly assessing the lockdown’s effect without accounting for temporal trends or interactions. 
   
The lower Adjusted R-squared values across all models suggest that other unaccounted factors might also influence changes in the prisoner population, and the presence of noise within the data. Variables such as policy changes, release rates, or crime trends could play a role.


## References

Ministry of Justice, Offender Management Statistics Quarterly publications, "Prison population: 30 June 2023", Retrieved on November 8, 2024, Available at: https://www.gov.uk/government/collections/offender-management-statistics-quarterly

