---
title: "Correlation Between Significant Events and Prison Population"
author: "by Gomakashi: Marcel, Sky, Hanna, Vivi, Siddhi & Maggie"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
library(ggplot2)

#PACKAGES YOU MIGHT NEED TO INSTALL
library(readODS)
library(janitor) #for changing row names
library(RColorBrewer)

```


```{r load-data, include=FALSE}
Ethnic_PD <- read_ods('data/Population_30June2023_Annual.ods', sheet = 14)

#wrapper function for generating title of plots
wrapper <- function(x, ...) 
{
  paste(strwrap(x, ...), collapse = '\n')
}

#normalises data to range from 0 to 1
normalise_01 <- function(x)
{
  if (min(x) < 0 )
  {
    x/max(abs(x))
  } else {
    (x-min(x))/(max(x)-min(x))
  }
}


```


## Research Question: Correlation between significant events and prison population of England and Wales

We have picked some major worldwide events to see whether if there is a noticeable correlation between them and changes in prison population.

## Data

We explored the 'Prison population by type of custody, age group and sex' section of the data.

The data set we looked at contained many custody groups like remand(individuals that are held in prison until their trial/hearing), intermediate sentence(individuals that have to stay in custody until they have been approved to be safe for the public), fine defaulter(individuals that are in prison due to refusing to pay a fine) and recalls(individuals that were released on parole/license but broke the rules of their probation and are back in prison). 

Note to future editor:
Elaborate in this section on:
- the time resolution (how much time is spaced between consecutive data), how it went from once a year to once a month in 2011, etc.)
- explaination of other data set used

## Findings

For the prime minister section, we included the whole data set and all the years to provide a comparison for before and during each Prime Ministers' time in office. There were a few limitations such as being unable to directly correlate causes of changes to specific polices that the government made at that time which resulted in assumptions being made. To help with this, we used different types of crime and looked into individual polices regarding those.

We then looked into specific factors and possible related government responses for both Lord Cameron and Theresa May:

For lord Cameron:
In Drug Offenses
The trend: After peaking around 2010, the prison population for drug offenses declined significantly during the Cameron years.
Key Issues:
- Reflects a shift in policing priorities and sentencing for drug-related crimes.
- Debate over the effectiveness of punitive measures versus rehabilitation programs.
Factors:
- Policing Strategy: Focus shifted to tackling high-level drug trafficking rather than low-level offenses.
- Rehabilitation: Increased emphasis on drug treatment programs as alternatives to imprisonment.
- Cultural Changes: Reduced stigma around certain drugs, particularly cannabis, influenced public and legal attitudes.
Government Response:
- Expanded funding for drug treatment and harm reduction initiatives.
- Focused on dismantling organized drug networks, deprioritizing minor possession cases.

For Fraud and Forgery
The trend: The prison population for fraud and forgery offenses declined steadily during this period.
Key Issues:
- Fewer resources allocated to investigating white-collar crime.
- Fraud increasingly perpetrated through cyber channels, making traditional arrests less feasible.
Factors:
- Technological Advances: Transition to online fraud reduced physical evidence, complicating arrests and convictions.
- Austerity: Police funding cuts impacted the capacity to investigate complex financial crimes.
- Focus Shift: Law enforcement prioritized violent and drug-related crimes over fraud.
Government Response:
- Encouraged the private sector to take more responsibility for preventing financial fraud.
- Supported the development of specialized units for cybercrime investigation.

For Motoring Offenses
The trend: The prison population for motoring offenses consistently declined throughout this period.
Key Issues:
- Declining imprisonment for motoring-related crimes could reflect a focus on alternative penalties.
- Technological enforcement (e.g., speed cameras) played a larger role than human intervention.
Factors:
- Automation: Greater use of automated systems reduced reliance on arrests for minor offenses.
- Legislation: Introduction of more on-the-spot fines and penalty point systems decreased incarceration rates.
- Austerity: Police budget cuts impacted road traffic enforcement.
Government Response:
- Focused resources on education campaigns (e.g., drink-driving awareness).
Promoted use of technology for road safety enforcement.

For Rape
The trend: The prison population for rape offenses steadily increased throughout the period.
Key Issues:
- Increased reporting due to awareness campaigns and social movements.
- Lengthier sentences for sexual offenses reflected public demand for justice.
Factors:
- Cultural Shift: Greater willingness of survivors to report sexual offenses.
- Judicial Changes: Harsher sentencing policies for rape and sexual violence.
- Awareness Campaigns: Initiatives like #MeToo raised awareness and led to action on sexual crimes.
Government Response:
- Expanded funding for sexual violence support services.
- Introduced training for police and prosecutors to handle rape cases sensitively.

For Robbery
The trend: The prison population for robbery offenses fluctuated, peaking around 2013 and declining thereafter.
Key Issues:
- Variability linked to changes in socioeconomic conditions and crime opportunities.
- Shift to non-violent theft methods like cybercrime reduced traditional robbery.
Factors:
- Economic Climate: Recession-related unemployment initially drove robbery rates up.
- Security Measures: Increased CCTV coverage deterred street-level robbery.
- Policing Priorities: Focus shifted to violent and high-profile crimes.
Government Response:
- Promoted urban safety initiatives, such as better street lighting.
- Strengthened penalties for armed robbery to deter violent crimes.

For Violence Against the Person
The trend: The prison population for violence against the person rose gradually until 2013, then fell off slightly.
Key Issues:
- Reflects stricter sentencing and prioritization of violent crime.
- Public concern about knife crime and domestic violence influenced policy focus.
Factors:
- Legislative Changes: Tougher sentences for assault and domestic violence.
- Social Awareness: Greater focus on addressing domestic violence and hate crimes.
- Policing Strategies: Proactive measures like stop-and-search increased violent crime detections.
Government Response:
- Introduced domestic violence legislation and support programs.
- Increased funding for community-based violence prevention initiatives.

For Violence Against The Person
The trend: Violent crime, including domestic violence, increased during Theresa May's time as Prime Minister. Austerity measures reduced police resources, but awareness campaigns led to more reporting of such crimes.
Key Issues:
- Rising arrests for violent crimes raised concerns about public safety.
- Domestic violence and knife crime were particularly alarming during her leadership.
Factors:
Increased Reporting: Campaigns encouraged victims to come forward.
Austerity: Budget cuts reduced the number of police officers.
Social Media: Highlighted incidents and raised public awareness.
Government Response:
- Introduced the Serious Violence Strategy to tackle knife crime and gang violence.
- Focused on addressing domestic violence with funding for shelters and new laws.

For Burglary 
The trend: Burglary rates declined during Theresa May’s tenure, partly due to improved home security systems. However, police budget cuts under austerity raised concerns about reduced capacity to address these crimes.
Key Issues :
- Decline in arrests may reflect reduced police resources rather than fewer burglaries.
- Continued reliance on homeowners to secure their properties.
Factors :
- Better Security: Increased use of CCTV, alarms, and smart locks deterred burglars.
- Austerity: Police funding cuts led to fewer officers investigating burglaries.
- Criminal Shift: Some offenders moved toward cybercrime instead of traditional burglary.
Government Response :
- Promoted neighborhood watch programs and public campaigns on home security.
- Focused police efforts on higher-priority crimes, leaving burglary less emphasized.

For Fraud and Forgery
The trend: Fraud increasingly shifted online during May’s tenure, with cybercrime emerging as a dominant concern. Traditional forgery and fraud arrests declined as enforcement adapted to these changes.
Key Issues:
- Cybercrime posed new challenges for law enforcement, requiring specialized skills and resources.
Factors:
- Technological Shift: Rapid digitization created opportunities for online fraud.
- Public Awareness: Many individuals and businesses lacked awareness of cybersecurity risks.
- Resource Limitations: Austerity limited the expansion of specialized cybercrime units.
Government Response:
- Expanded the National Crime Agency’s cybercrime capabilities.
- Promoted public education campaigns to raise awareness of online fraud risks.

For Drug Offences
The trend: Drug-related crime remained a persistent issue during May’s tenure, with a focus on tackling organized crime networks and trafficking rather than low-level offenses.
Key Issues:
- A fluctuating trend in drug-related arrests reflected changing enforcement strategies.
- Rising concern over county lines drug operations involving vulnerable individuals.
Factors:
- Organized Crime: Increased trafficking and exploitation of minors in drug distribution.
- Reduced Policing Capacity: Austerity limited proactive drug enforcement efforts.
- Changing Social Attitudes: Shifts in societal views on recreational drug use affected enforcement priorities.
Government Response:
- Strengthened the role of the National Crime Agency (NCA) in combating drug trafficking.
- Launched initiatives to address county lines drug operations, including safeguarding vulnerable individuals.

For Rape
The trend: Sexual violence reporting increased during May’s tenure due to heightened awareness and improved survivor support systems. However, there was criticism about the low prosecution rates for these crimes.
Key Issues:
- Rising arrests reflected improved reporting but also revealed gaps in handling sexual violence cases.
Factors:
- Social Movements: Global campaigns like #MeToo increased awareness and reporting of sexual violence.
- Police Training: Law enforcement became better equipped to handle sensitive cases.
- Judicial Backlogs: Delays in prosecution and low conviction rates caused frustration among victims
Government Reponse:
- Promoted better police training on handling sexual offenses.
- Advocated for increased funding for victim support programs and rape crisis centers.
- Supported the creation of specialist units to handle sexual violence cases.

For Robbery
The Trend: Fluctuations in robbery arrests reflected targeted interventions in high-crime areas, though austerity measures hindered broader prevention efforts.
Key Issues:
- Robbery remained a concern, especially in urban areas with higher youth crime rates.
Factors:
- Youth Involvement: Economic challenges during austerity may have driven some young people toward robbery.
- Targeted Policing: Police focused resources on areas with high robbery rates.
- Technological Advances: Improved surveillance helped deter robberies.
Government Response:
- Promoted community engagement programs to prevent youth involvement in robbery.
- Supported the use of data-driven policing to identify and target robbery hotspots.

For Other Offenses
The Trend: An increase in arrests for miscellaneous crimes, such as public order offenses and emerging cyber-related crimes, highlighted changing priorities in law enforcement under May’s leadership.
Key Issues:
- Law enforcement struggled to address modern challenges with reduced resources.
Factors:
- Emerging Crimes: Growth in cyber-related offenses and technology-driven crimes.
- Shifting Social Norms: Changes in public behavior led to new categories of offenses.
Government Response:
- Modernized police training to address evolving threats, such as cybercrime and organized crime.
- Introduced the Modern Slavery Act 2015, targeting exploitation and human trafficking.

For Motoring Offenses
The Trend: During Theresa May’s time, arrests for motoring offenses dropped due to increased use of automated systems like speed cameras, reducing the need for police intervention.
Key Issues:
- Fewer arrests raised concerns about gaps in addressing reckless driving.
- Police presence on roads decreased due to budget cuts.
Factors:
- Technology: Automated tools like speed cameras took over traffic enforcement.
- Awareness: Road safety campaigns improved driver behavior.
- Austerity: Police had fewer resources for traffic enforcement.
Government Response:
- Supported automated traffic systems.
- Promoted road safety campaigns.
- Prioritized police resources for more urgent crimes.


The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.




## References

Ministry of Justice, Offender Management Statistics Quarterly publications, "Prison population: 30 June 2023", Retrieved November 8, 2024, Available at: https://www.gov.uk/government/collections/offender-management-statistics-quarterly

