---
title: "Correlation Between Significant Events and Prison Population"
author: "by Gomakashi: Marcel, Sky, Hanna, Vivi, Siddhi & Maggie"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
library(ggplot2)

#PACKAGES YOU MIGHT NEED TO INSTALL
library(readODS)
library(janitor) #for changing row names
library(RColorBrewer)

```


```{r load-data, include=FALSE}
Ethnic_PD <- read_ods('data/Population_30June2023_Annual.ods', sheet = 14)

#wrapper function for generating title of plots
wrapper <- function(x, ...) 
{
  paste(strwrap(x, ...), collapse = '\n')
}

#normalises data to range from 0 to 1
normalise_01 <- function(x)
{
  if (min(x) < 0 )
  {
    x/max(abs(x))
  } else {
    (x-min(x))/(max(x)-min(x))
  }
}


```


## Research Question: Correlation between significant events and prison population of England and Wales

We have picked some major worldwide events to see whether if there is a noticeable correlation between them and changes in prison population.

## Data

We explored the 'Prison population by type of custody, age group and sex' section of the data.

The data set we looked at contained many custody groups like remand(individuals that are held in prison until their trial/hearing), intermediate sentence(individuals that have to stay in custody until they have been approved to be safe for the public), fine defaulter(individuals that are in prison due to refusing to pay a fine) and recalls(individuals that were released on parole/license but broke the rules of their probation and are back in prison). 

Note to future editor:
Elaborate in this section on:
- the time resolution (how much time is spaced between consecutive data), how it went from once a year to once a month in 2011, etc.)
- explaination of other data set used

## Findings

For the prime minister section, we included the whole data set and all the years to provide a comparison for before and during each Prime Ministers' time in office. There were a few limitations such as being unable to directly correlate causes of changes to specific polices that the government made at that time which resulted in assumptions being made. To help with this, we used different types of crime and looked into individual polices regarding those.

The write-up of your project and findings go here. 

Think of this as the text of your presentation with some extra detail to cover what there was not time to discuss in the presentation. This might include any assumptions you made when doing your analysis, any limitations of the work you have done or any ideas you have for future work. Feel free to split this section into subsections to make it easier to read. 

The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.





## References

Ministry of Justice, Offender Management Statistics Quarterly publications, "Prison population: 30 June 2023", Retrieved November 8, 2024, Available at: https://www.gov.uk/government/collections/offender-management-statistics-quarterly

