---
title: "Correlation Between Significant Events and Prison Population"
author: "by Gomakashi: Marcel, Sky, Hanna, Vivi, Siddhi & Maggie"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
library(ggplot2)

#PACKAGES YOU MIGHT NEED TO INSTALL
library(readODS)
library(janitor) #for changing row names
library(RColorBrewer)

```


```{r load-data, include=FALSE}
Ethnic_PD <- read_ods('data/Population_30June2023_Annual.ods', sheet = 14)

#wrapper function for generating title of plots
wrapper <- function(x, ...) 
{
  paste(strwrap(x, ...), collapse = '\n')
}

#normalises data to range from 0 to 1
normalise_01 <- function(x)
{
  if (min(x) < 0 )
  {
    x/max(abs(x))
  } else {
    (x-min(x))/(max(x)-min(x))
  }
}


```


## Research Question: Correlation between significant events and prison population of England and Wales

We have picked some major worldwide events to see whether if there is a noticeable correlation between them and changes in prison population.

## Data

We explored the 'Prison population by type of custody, age group and sex' section of the data.

The data set we looked at contained many custody groups like remand(individuals that are held in prison until their trial/hearing), intermediate sentence(individuals that have to stay in custody until they have been approved to be safe for the public), fine defaulter(individuals that are in prison due to refusing to pay a fine) and recalls(individuals that were released on parole/license but broke the rules of their probation and are back in prison). 

Note to future editor:
Elaborate in this section on:
- the time resolution (how much time is spaced between consecutive data), how it went from once a year to once a month in 2011, etc.)
- explaination of other data set used

## Findings
### Prime Minister
For the prime minister section, we included the whole data set and all the years to provide a comparison for before and during each Prime Ministers' time in office. There were a few limitations such as being unable to directly correlate causes of changes to specific polices that the government made at that time which resulted in assumptions being made. To help with this, we used different types of crime and looked into individual polices regarding those.

We then looked into specific factors and possible related government responses for both Lord Cameron and Theresa May.

#### Lord Cameron(2010-2016):

##### 1.Drug Offences

Trend: 
After peaking around 2010, the prison population for drug offences declined significantly during the Cameron years.

Key Issues:

- Reflects a shift in policing priorities and sentencing for drug-related crimes.

- Debate over the effectiveness of punitive measures versus rehabilitation programs.

Factors:

- Policing Strategy: Focus shifted to tackling high-level drug trafficking rather than low-level offenses.

- Rehabilitation: Increased emphasis on drug treatment programs as alternatives to imprisonment.

- Cultural Changes: Reduced stigma around certain drugs, particularly cannabis, influenced public and legal
attitudes.

Government Response:

- Expanded funding for drug treatment and harm reduction initiatives.

- Focused on dismantling organized drug networks, deprioritizing minor possession cases.


##### 2.Fraud Offences

Trend: 
Fraud offenses steadily declined during Cameron’s tenure, reflecting reduced resources for white-collar crime investigations and a shift toward online fraud.

Key Issues:

- Fewer resources allocated to investigating white-collar crime.

- Fraud increasingly perpetrated through cyber channels, making traditional arrests less feasible.

Factors:

- Technological Advances: Transition to online fraud reduced physical evidence, complicating arrests and convictions.

- Austerity: Police funding cuts impacted the capacity to investigate complex financial crimes.

- Focus Shift: Law enforcement prioritized violent and drug-related crimes over fraud.

Government Response:

- Encouraged the private sector to take more responsibility for preventing financial fraud.

- Supported the development of specialized units for cybercrime investigation.


##### 3.Rape

Trend: 
The prison population for rape offenses steadily increased throughout the period.

Key Issues:

- Increased reporting due to awareness campaigns and social movements.

- Lengthier sentences for sexual offenses reflected public demand for justice.

Factors:

- Cultural Shift: Greater willingness of survivors to report sexual offenses.

- Judicial Changes: Harsher sentencing policies for rape and sexual violence.

- Awareness Campaigns: Initiatives like #MeToo raised awareness and led to action on sexual crimes.

Government Response:

- Expanded funding for sexual violence support services.

- Introduced training for police and prosecutors to handle rape cases sensitively.


##### 4.Robbery

Trend: 
Robbery offenses peaked early in Cameron’s tenure and then declined significantly, reflecting improvements in urban security and targeted policing.

Key Issues:

- Variability linked to changes in socioeconomic conditions and crime opportunities.

- Shift to non-violent theft methods like cybercrime reduced traditional robbery.

Factors:

- Economic Climate: Recession-related unemployment initially drove robbery rates up.

- Security Measures: Increased CCTV coverage deterred street-level robbery.

- Policing Priorities: Focus shifted to violent and high-profile crimes.

Government Response:

- Promoted urban safety initiatives, such as better street lighting.

- Strengthened penalties for armed robbery to deter violent crimes.


##### 5.Theft Offences 

Trend: 
Theft offenses saw an initial rise and then consistently declined, reflecting improved security and changing criminal behaviors.

Key Issues:

- Improved home and business security measures reduced opportunities.

- A shift from physical theft to cybercrime influenced trends.

Factors:

- Technological Advances: Alarms and surveillance reduced physical theft.

- Policing Priorities: Focused resources on higher-profile crimes.

Government Response:

- Encouraged neighborhood watch programs and public safety campaigns.

- Supported private-public collaborations to prevent theft.


#### Theresa May(2016-2019):

##### 1.Drug Offences

Trend:
Drug offences saw fluctuations, with an initial decline followed by a rise during May's leadership, reflecting shifts in enforcement strategies.

Key Issues:

- Increase in county lines drug operations exploiting vulnerable individuals.

- Resource constraints limited comprehensive enforcement.

Factors:

- Organized Crime: Rise in county lines operations.

- Austerity: Reduced police budgets hindered proactive measures.

Government Response:

- Strengthened the National Crime Agency (NCA) to tackle drug trafficking.

- Introduced targeted initiatives to safeguard vulnerable individuals.


##### 2.Fraud Offenses

Trend: 
Fraud offenses steadily declined, partly due to a shift toward cyber fraud, which is harder to detect and prosecute.

Key Issues:

- Limited resources for investigating complex financial crimes.

- Increasing prevalence of online scams.

Factors:

- Technology: Shift to online fraud reduced physical evidence.

- Austerity: Cutbacks impacted investigative capacity.

Government Response:

- Promoted private sector collaboration in fraud prevention.

- Strengthened cybercrime units within the NCA.


##### 3.Rape

Trend: 
Rape offences rose sharply, reflecting increased awareness and improved reporting mechanisms during May’s tenure.

Key Issues:

- Growing public awareness of sexual violence.

- Low conviction rates despite rising reports.

Factors:

- Cultural Shift: Movements like #MeToo encouraged reporting.

- Judicial Reforms: Harsher sentencing for sexual crimes.

Government Response:

- Expanded victim support services.

- Improved police training for handling sensitive cases.


##### 4.Robbery

Trend: 
Robbery offenses saw a steep decline, likely due to increased use of surveillance and targeted interventions in urban areas.

Key Issues:

- Economic challenges initially drove up robbery rates.

- Policing focus shifted to violent and organized crime.

Factors:

- Security Measures: Increased CCTV and street lighting.

- Economic Conditions: Recession impacts reduced over time.

Government Response:

- Promoted urban safety initiatives.

- Strengthened penalties for violent robberies.


##### 5.Theft Offences

Trend: 
Theft offences consistently declined, driven by improved security measures and technological advancements.

Key Issues:

- Shift from physical theft to online fraud and scams.

- Declining police presence in rural areas.

Factors:

- Technology: Use of alarms and surveillance deterred theft.

- Policing Priorities: Resources focused on violent crimes.

Government Response:

- Encouraged public safety campaigns.

- Supported neighborhood watch programs to reduce theft.









The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.




## References

Ministry of Justice, Offender Management Statistics Quarterly publications, "Prison population: 30 June 2023", Retrieved November 8, 2024, Available at: https://www.gov.uk/government/collections/offender-management-statistics-quarterly

